Estimating mismapping rates in seed-based algorithms for read mapping

### 2019-03-03 MEM.tex
Adding theory for skip seeds and exact seeds.

\usepackage{algorithm2e}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{blkarray}
\usepackage[font=small]{caption}
\usepackage{cite}
\usepackage{graphicx}

@@

\newcommand{\dn}[1]{\scriptstyle{\downarrow_{/#1}}}
\newcommand{\up}[1]{\scriptstyle{\uparrow_{/#1}}}
\newcommand{\nd}{\scriptstyle{|}}

@@

\title{Computing MEM seeding probabilities}

@@

\begin{abstract}
The increasing throughput of DNA sequencing technologies demands faster
algorithms to analyze sequencing data. The fate of most reads is to be
mapped to a reference sequence, typically a genome. Modern mappers rely on
heuristics to gain speed at a reasonable cost for accuracy. MEM seeding
(where MEM stands for Maximal Exact Match) is a popular heuristic that
shows good empirical performance but for which theory is lacking. It is
thus difficult to calibrate or optimize mapping strategies based on MEM
seeding. Here we develop a theory to estimate the probability that the MEM
seeding heuristic fails to return the correct candidate. The major
innovation is to use concepts from analytic combinatorics to represent
reads as abstract sequences, from which we estimate the probabilities of
interest. We provide two separate methods, which combined together give a
practicial solution to the problem of calibrating MEM seeding when
aligning short reads. These results have implications for MEM-based
heuristics and lay the foundation of a general strategy to tackle sequence
alignment problems.
\end{abstract}

@@

We will refer to this task as the \emph{best location problem}. It amounts
to finding the optimal alignment between two sequences, and for this
reason has received substantial attention in bioinformatics. It turns out
that finding the best location of the read is the only way to get to its
true location, so for now we need to focus our attention on the best
location problem.

@@

There exists exact algorithms to solve the best location
problem~\cite{pmid7265238,pmid5420325}, but they are too slow to process
the large amount of data generated by modern sequencers. Instead, one uses
heuristic methods, \textit{i.e.} algorithms that run fast, but may not
return the best location~\cite{Waterman1984}. Heuristic mapping algorithms
thus have three possible outcomes: $i.$ the DNA fragment is mapped to the
best location, $ii.$ the DNA fragment is mapped to some other location,
$iii.$ the DNA fragment is not mapped at all.

The most popular heuristic for this problem is to first identify short
regions of high similiarity between the read and the genome, and then run
an exact alignment algorithm locally at those regions. The first step is a
filtration called ``seeding'' and the regions of high similarity are
called ``seeds''. This is for instance the general strategy of the popular
alignment algorithm BLAST~\cite{pmid2231712}.

If seeds can be discovered fast, we can quickly zoom in on a small set of
candidate locations and reduce the input size of the (slow) alignment
algorithm. The disadvantage is that the target location is not always in
the candidate set. In this case, it cannot be discovered and the outcome
is either $ii.$ or $iii.$, which are both erroneous.

The slowness of the alignment algorithm imposes a trade-off between speed
and accuracy. If the seeding step is set to produce many candidates, the
target is likely to be among them and it will be discovered. The downside
is that this will take long. Conversely, if the seeding step is set to
produce few candidates, the process will run faster but the target is less
likely to be in the candidate list so it is less likely to be discovered.

Importantly, the terms of this trade-off depend on the algorithm. Some
seeding methods are better than others (see~\cite{pmid20460430,
sun2005designing,pmid11934743,xu2006optimizing,pmid17044164,
brejova2003vector,pmid18684737,pmid15359419} for examples of design,
see~\cite{pmid16533404, pmid20460430} for comparisons and discussions, and
see~\cite{navarro2001guided} for a more general review of filtration
methods in pattern matching). The upshot is that we can develop faster
mapping algorithms at no cost for accuracy if we find better seeding
strategies.

The seeding framework was developed for the best location problem, but it
also applies to the true location problem. Actually, the process is
exactly the same except that outcomes $i.$ and $ii.$ now refer to the true
location. In outcome $i.$ called a ``true positive'', the DNA fragment is
mapped to the \emph{true} location. In outcome $ii.$ called a ``false
positive'', the DNA fragment is mapped to an incorrect location. Outcome
$iii.$ called a ``false negative'' is the same as before and means that
the DNA fragment is not mapped at all.

We will make two assumptions that will simplify the discussion. The first
is that the true location is always the best. We will further comment on
this and explain how to relax this assumption in section
section~\ref{sec:not_the_best}. Until then, we consider that the true
location problem is the same as the best location problem. The second
assumption is that the best location is always discovered if it is in the
candidate set. This implies that outcomes $i.$, $ii.$ and $iii.$ are
determined by the seeding process. For this reason, we will sometimes
refer to them as the outcomes of the seeding step, even though strictly
speaking they are the outcomes of the whole mapping pipeline\footnote{This
can be confusing because some authors refer to the off-target seeds as
false positives. When we say that the outcome of the seeding step is a
false positive, we mean that \emph{all} the seeds are off-target.}.

@@

In general, seeds do not have to be exact matches{}. For
instance, the algorithm PatternHunter~\cite{pmid11934743} uses ``spaced
seeds'' that tolerate mismatches\footnote{For the anecdote, the method is
covered by a patent (reference US20070088510A1).}. However, in this
document, we will always assume that a seed is an exact match with no gap.
In other words, a seed is a perfect match between a part of the read and a
genomic site.

The rest of the document will focus on a seeding heuristic known as ``MEM
seeding'' (where MEM stands for Maximal Exact Match). This approach gives
good empirical results when mapping short reads (it is for instance the
strategy\footnote{To be exact, BWA-MEM uses SMEM seeds (Super Maximal
Exact Match seeds).} of the popular mapper BWA-MEM~\cite{li2013aligning})
but there is presently no theory to justify its performance. A deeper
understanding of MEM seeding would be useful to calibrate the
speed/accuracy trade-off of this heurisitic.

@@

This definition presents a computational challenge: to know if a given
subsequence is a MEM seed we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately there
exists practical methods to identify MEM seeds, even when the reference
genome is very large~\cite{ferragina2000opportunistic,
ferragina2005indexing}. Such algorithms are crucial for the present
discussion, but describing them falls outside the scope of the document.
We refer the interested reader to the relevant
literature~\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736,
li2013aligning}, and here on we will assume that MEM seeds are available
at all times. Looking ahead to section~\ref{sec:est}, let us just mention
that this can be done with an algorithm called the \emph{backward
search}~\cite{ferragina2000opportunistic}.

>>>>

\usepackage{algorithm2e}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{blkarray}
\usepackage[font=small]{caption}
\usepackage{cite}
\usepackage{enumerate}
\usepackage{graphicx}

@@

\newcommand{\dn}[1]{\scriptstyle{\downarrow_{/#1}}}
\newcommand{\Dn}[1]{\scriptstyle{\Downarrow_{#1}}}
\newcommand{\up}[1]{\scriptstyle{\uparrow_{/#1}}}
\newcommand{\nd}{\scriptstyle{|}}
\newcommand{\modulo}[1]{\ (\mathrm{mod}\ #1)}

@@

\title{Estimating mismapping rates in seed-based algorithms
for read mapping}

@@

\begin{abstract}
% Statement of the problem.
The increasing throughput of DNA sequencing technologies demands faster
algorithms to analyze sequencing data. The fate of most reads is to be
mapped to a reference sequence, typically a genome. Modern mappers rely on
heuristics to gain speed at a reasonable cost for accuracy. In the seeding
approach, short matches between the reads and the genome are used to
narrow the search to a set of candidate locations. Several seeding
variants used in modern mappers show good empirical performance but
theoretical results are still lacking, making them difficult to calibrate
or top optimize.
% Summary of the results.
Here we develop a theory to estimate the probability that reads are
mismapped due to a seeding failure. We describe the properties of simple
exact seeds, skip-seeds used in Bowtie2 and MEM (Maximal Exact Match)
seeds used in BWA-MEM. 
% Innovation.
The major innovation is to use concepts from analytic combinatorics to
represent reads as abstract sequences, from which we estimate the
probabilities of interest.
% Durability.
We provide several algorithms, which combined together give a practicial
solution to the problem of calibrating seeding heuristic for short reads.
% Scope.
These results significantly improve current mapping algorithms and lay the
foundation of a general strategy to tackle sequence alignment problems.
\end{abstract}

@@

We will refer to this task as the \emph{best location problem}. It amounts
to finding the optimal alignment between two sequences, and for this
reason has received substantial attention in bioinformatics. For
simplicity we will assume that the true location is always the best. We
will relax this assumption in section section~\ref{sec:not_the_best}, but
until then we will consider that the solution of the true location problem
is the best location.

@@

There exists algorithms to solve the best location
problem~\cite{pmid7265238,pmid5420325}, but they are too slow to process
the large amount of data generated by modern sequencers. Instead, one uses
heuristic methods, \textit{i.e.} algorithms that run fast, but that may
return an incorrect result~\cite{Waterman1984}.

The most popular heuristic for the best location problem is a filtration
method called ``seeding''. The principle is to first identify short
regions of high similiarity between the read and the genome, and then run
an exact alignment algorithm at the candidate locations. This is for
instance the general strategy of the popular alignment algorithm
BLAST~\cite{pmid2231712}.

If seeding is fast, we can quickly zoom in on a small set of candidate
locations and reduce the input size of the (slow) alignment algorithm. The
disadvantage is that the target location is not always in the candidate
set. When that happens, the read cannot be mapped to the correct location
because it has been filtered out at the seeding step.

The slowness of the alignment algorithm imposes a trade-off between speed
and accuracy: If the seeding step is set to produce many candidates, it is
likely to retrieve the candidate. The downside is that this will take
long. Conversely, if the seeding step is set to produce few candidates,
the process will run faster but the target is less likely to be in the
candidate list.

Importantly, the trade-off depends on the seeding method. This means that
we can develop faster mapping algorithms at no cost for accuracy, as long
as we can find better seeding strategies. Progress on this line of
research has largely benefited from the improvement of computer hardware
and from the deveopment of indexing algorithms. Literature on this topic
is already abudant. Some examples of seeding algorithms are given in
references~\cite{sun2005designing,pmid11934743,xu2006optimizing,
pmid17044164,brejova2003vector,pmid18684737,pmid15359419}.
References~\cite{pmid16533404} and~\cite{pmid20460430} present high-level
comparisons of different designs, and reference~\cite{navarro2001guided}
gives a global overview of filtration methods in pattern matching.

\subsection{The two types of seeding failure}

Filtering methods are considered to fail whenever the target is not in the
candidate set. For seeding heuristics, we need to distinguish two
different kinds of failure. In the first kind, the candidate set contains 
a duplicate of the target but does not contain the target itself; in the
second kind, the candidate set contains neither.

The distinction is important because duplicates of the target are similar
to the read (due to their similarity to the target), so a failure of the
first kind is difficult to distinguish from a success. A failure of the
second kind is easier to recognize because in this case the best hit is
not similar to the read.

Before going further, we introduce some vocabulary that will be useful to
simplify the exposition of the theory.

\begin{definition}
The seeding process is said to be
\begin{enumerate}[i)]
\item on target if the candidate set contains the target,
\item off target if the candidate set contains a duplicate but
not the target, 
\item null if the candidate set contains neither.
\end{enumerate}
\end{definition}

If a seed-based algorithm returns an incorrect location, then the best
location was never tested with an exact alignment. If the error is given a
high confidence, then it must be the best in the candidate set and it must
be similar to the read. This narrows down the cases of mismapping, or more
exactly high confidence mismapping to off-target seeding.

This observation is fundamental. In order to estimate the mismapping rate
of an algorithm, we must know how often the associated seeding heuristic
is off target. Yet, this information is missing for nearly all seeding
methods, meaning that current mappers cannot be properly calibrated.

The focus of this article is to develop a theory to estimate the
probability of off-target seeding under the most popular seeding
heuristics. Our previous work pioneered a method to compute seeding
probabilities but it did not distinguish off-target from null
seeding~\cite{filion2017analytic,filion2018analytic}, and therefore did
not provide a way to control the mismapping rate. Other authors
investigated the reliability of mapping algorithms~\cite{pmid23872968},
but they focused on the probability that random sequences have a single
hit, recognizing that solving the problem requires taking into
consideration the repeat structure of the genome.

\section{Types of seeds}

The term ``seed'' does not have exactly the same meaning according to
different authors. It can designate a part of the read, a part of the
genome, a particular sequence motif or a structured pattern of matches.
Also, the term is not always short for ``exact match''. For instance, the
algorithm PatternHunter~\cite{pmid11934743} describes ``spaced seeds''
that tolerate mismatches\footnote{For the anecdote, the method is covered
by a patent (reference US20070088510A1).}. To avoid potential confusion,
we will adopt the following convention.

\begin{definition}
A seed is a subsequence of the read that is above a certain size $\gamma$
(defined by the context of the problem) and that has at least one perfect
match in the reference genome.
\end{definition}

When a seed matches a particular location of the genome, we will say that
it is a seed \emph{for} that location. For instance, we will often refer
to a seed for the target or a seed for one of its duplicates.

This definition presents a computational challenge: to know if a given
subsequence is a seed we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately there
exists practical methods to solve it, even when the reference genome is
very large~\cite{ferragina2000opportunistic, ferragina2005indexing}. Such
algorithms are crucial for the present theory, but describing them
falls outside the scope of the document. Looking ahead to
section~\ref{sec:est}, let us only mention that this can be done with an
algorithm called the \emph{backward
search}~\cite{ferragina2000opportunistic}. We refer the interested reader
to the relevant
literature~\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736} and
here on we will assume that seeds are always known without ambiguity [ALL
THE CITED LITERATURE IS FOR MEM SEEDS, WE NEED TO FIND ARTICLES THAT TREAT
THE CASE OF OTHER SEEDS].

\subsection{Exact seeds}

Exact seeds have a fixed size $\gamma$. When using exact seeds, all the
perfect matches of size $\gamma$ are collected to constitute the candidate
set. This seeding heuristic was used in the first version of
BLASTN~\cite{pmid2231712}, but is no longer used because they produce many
overlapping matches in the same regions, as we are about to illustrate.

Fig.~\ref{fig:exact_seed_example} shows the example of a read produced
by an imperfect sequencer. The sequenced DNA fragment comes from a region
of the genome that has three extra duplicates, and the read has three
miscalled nucleotides (indicated by stars).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{exact_seed_example.pdf}
\caption{\textbf{Example of exact seeds.}
A sequenced DNA fragment (Read) is shown above its original form in the
genome (Target).}
\label{fig:exact_seed_example}
\end{figure}

Fig.~\ref{fig:exact_seed_example} also illustrates the seeding process
with exact seeds of size $\gamma=7$. It is clear that erroneous
nucleotides can also belong to exact seeds because they sometimes match
one of the duplicates. However, this is not the case of erroneous
nucleotides that are also mismatches for \emph{all} the duplicates, like
the second sequencing error in this example.

Thie middle of the read shows some cluttering where consecutive seeds
match consecutive sequences at the same location. This is typical for
exact seeds and constitutes a nuisance for the implementation. Indeed, it
is a waste of computer resources to discover matches in sequences that are
already in the candidate set. In addition, this seeding method is not
particularly sensitive~\cite{pmid11934743} so it is used only for specific
applications. Nevertheless, it will be extremely useful for the
development of the present theory.

\subsection{Skip seeds}

Skip seeds have a fixed size $\gamma$, but unlike exact seeds they cannot
start at every nucleotide. Instead, a fixed amount of nucleotides is
skipped between every seed. This is a way to reduce the amount of
overlapping matches at the same location, but the cost is a loss of
sensitivity. This seeding heuristic is used by
Bowtie2~\cite{pmid22388286}, where seeds have size $\gamma=16$ and fall
every 10 nucleotides (9 nucleotides are skipped between seeds).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{skip_seed_example.pdf}
\caption{\textbf{Example of skip seeds.}
The sequences and the annotations are the same as in
Fig.~\ref{fig:exact_seed_example}, but here one nucleotide is skipped
betwen the seeds. In other words, seeds can never start on a nucleotide
with an even index. To highlight the difference with
Fig.~\ref{fig:exact_seed_example}, the missing seeds are represented by
dotted lines. Skipping nucleotides reduces the clutter at the center of
the read, but it also removes the only seed for the second duplicate.}
\label{fig:skip_seed_example}
\end{figure}

Fig.~\ref{fig:skip_seed_example} shows the seeding process on the read of
Fig.~\ref{fig:exact_seed_example} when using skip seeds instead of exact
seeds. Here the size is still $\gamma=7$ but 1 extra nucleotide is skipped
between seeds. This amounts to removing every other seed. The consequence
is that there are fewer overlapping matches at the center of the read, but
the only seed for the second duplicate was lost. This is a rather positive
outcome because there is one off-target location fewer in the candidate
set, but in general the same might happen to the target.

It is clear that skipping nucleotides reduces the sensitivity of the
seeding, but to what extent? One could test this empirically, but it
depends on the seed length, the amount of nucleotides that are skipped,
the error rate of the sequencer and the size of the reads. The present
theory will allow us to get the answer automatically.

@@

We will now focus on a seeding heuristic known as ``MEM seeding'' (where
MEM stands for Maximal Exact Match). This approach gives good empirical
results when mapping short reads (it is for instance the
strategy\footnote{To be exact, BWA-MEM uses SMEM seeds (Super Maximal
Exact Match seeds).} of the popular mapper BWA-MEM~\cite{li2013aligning})
but there is presently no theory to justify its performance. A deeper
understanding of MEM seeding would be useful to calibrate the
speed/accuracy trade-off of this heurisitic.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=1]{MEM_example.pdf}
\caption{\textbf{Example of MEM seeds.}
A DNA fragment is sequenced with an imperfect instrument. It comes from
the top genomic location (Truth), but three errors were introduced
(indicated by stars). The sequence of the DNA fragment has three
duplicates at other genomic locations, shown below the target. Mismatches
between the read and the candidate sequence are highlighted in bold. MEM
seeds are indicated by grey boxes.}
\label{fig:MEM_example}
\end{figure}

\section{Model and computational strategy}
\label{sec:symbolic}

\subsection{Sequencing error and duplications}

Now that this is clarified, we need to model the sequencing and the
duplication processes so that we can compute the probabilities of the
events of interest. We assume that the sequencing instrument has a
constant substitution rate $p$, and that insertions and deletions never
occur. When a substitution occurs, we assume that the instrument is
equally likely to output any one of the remaining three nucleotides. This
is a coarse-grain description of the error spectrum of the Illumina
technology~\cite{pmid21576222}.

Next, we assume that the target sequence has $N \geq 0$ duplicates, so
that there are $N$ off-target sequences. We further assume that
duplication happened instantaenously at some point in the past and that
all $N+1$ sequences diverge independently of each other at a constant
rate. In other words, we ignore the complications due to the genealogy of
the duplication events and we simply assume that at each position, any
given duplicate is identical to the target with probability $1-\mu$. If it
is not, we assume that the duplicate has any of the remaining three
nucleotides with equal probability (\textit{i.e.} each is found with
probability $\mu/3$).

Note that read errors are always mismatches against the target (because we
assume that the target is the true sequence), and they match each
off-target sequence with probability $\mu/3$. Correct nucleotides are
always a match for the target, and they match each off-target sequence
with probability $1-\mu$.

The central object of analytic combinatorics is the generating function,
and for our particular purpose we will use a special kind known as
\emph{weighted generating function}.

More didactic descriptions of the strategy to compute seeding
probabilities using analytic combinatorics have been published
elsewhere~\cite{filion2017analytic,filion2018analytic}. The interested
readers can also discover more about analytic combinatorics in the
standard textbooks~\cite{flajolet2009analytic,sedgewick2013introduction}.

\subsection{Example 1: on-target exact seeds}
\label{sec:sample_exact}

We highlight the strategy above with an example that will turn out to be
central for the development of the theory. In addition, it is simple
enough to provide a gentle introduction to the general methodology. This
example was described in detail in~\cite{filion2017analytic} and
in~\cite{filion2018analytic}, but we repeat it here with a formalism that
fits the present manuscript.

The first step is to note that the nucleotide sequence of the read is
irrelevant. Indeed, the read has an on-target seed if and only if it
contains a stretch of $\gamma$ nucleotides without error. So the main step
is to recode the read as sequence of correct or erroneous nucleotides.

We define the mismatch alphabet $\mathcal{A}_0 = \{\square, |,
\Downarrow,\}$, where $\square$ represents a correct nucleotide,
$\Downarrow$ represents an erroneous nucleotide and $|$ is special symbol
appended after the last nucleotide to mark the end of the read.

This recoding allows us to segment the read in \emph{segments} defined as
a sequence of 0 or more symbols $\square$ followed by a symbol
$\Downarrow$. The last part of the read is called the tail and it is
terminated by the symbol $|$ instead of $\Downarrow$. The symbol $|$ does
not correspond to a nucleotide, it has size 0. This decomposition is
unique, so we can view a read as a sequence of segments with a tail,
instead of a sequence of nucleotides.

Fig.~\ref{fig:simple} shows an example of decomposition in segments.
On-target seeds cannot contain sequencing errors, therefore they must lie
inside a segment or inside the tail. So the size of the segments and of
the tail indicates whether the read contains an on-target seed.

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_simple.pdf}
\caption{\textbf{The mismatch encoding.}
A example read is represented in the match alphabet. The symbol
$\Downarrow$ represents a mismatch against the target sequence and
the symbol $\square$ represents a match.}
\label{fig:simple}
\end{figure}

The probability of occurrence of the symbol $\Downarrow$ is $p$, the error
rate of the sequencer, and the probability of occurrence of the symbol
$\square$ is thus $1-p = q$. Both symbols have size 1, so their respective
weighted generating function are $pz$ and $qz$. Using the rule for
concatenation, we see that the weighted generating function of a segment
with $m$ symbols $\square$ is $(qz)^mpz$. The symbol $|$ has size 0, so
the weighted generating function of a tail segment with $m$ symbols
$\square$ is $(qz)^m$.

The key insight is that reads without on-target seed correspond exactly
with the reads that are made of segments and tail with fewer than $\gamma$
symbols $\square$. The weighted generating function of such segments is
$pz \cdot \big( 1+qz+\ldots+(qz)^{\gamma-1} \big)$, and that of the tail
is $1+qz+\ldots+(qz)^{\gamma-1}$. This gives a construction plan that can
be encoded as a transfer matrix.

Since there are two kinds of bjects, the segments and the tails, the
dimensions of the transfer matrix are $2 \times 2$. A segment can be
followed by another segment or by the tail. The tail can be followed by
nothing. The expression of the transfer matrix $M(z)$ is thus
\begin{equation*}
\begin{blockarray}{ccc}
   & \scriptstyle{\Downarrow} & \scriptstyle{|} \\
\begin{block}{c[cc]}
\scriptstyle{\Downarrow} & pz \cdot \big( 1+qz+\ldots+(qz)^{\gamma-1}
\big)  & 1+qz+\ldots+(qz)^{\gamma-1} \\
\scriptstyle{|} & 0 & 0 \\
\end{block}
\end{blockarray}.
\end{equation*}

In this case, it is possible to compute $M(z) + M(z)^2 + \ldots = M(z)
\cdot (I-M(z))^{-1}$ directly and to extract the entry of interest.
Here it is the top right term because the read can start with a segment or
the tail (as if the read were prepended by an invisible segment), and it
must end with the tail. It is easy to check that the weighted generating
function is equal to
\begin{equation}
\label{eq:F}
\frac{1+qz+\ldots+(qz)^{\gamma-1}}
  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)}.
\end{equation}

The function above can be represented as a Taylor series $a_0 + a_1z +
a_2z^2 + \ldots$ By construction, $a_k$ is the quantity of interest,
\textit{i.e.} the probability that a read of size $k$ does not contain an
on-target seed, so we now need to extract those coefficients from
expression (\ref{eq:F}). There are several methods to do so, the one we
choose here is to build a recurrence equation to compute $a_k$ from the
previous terms of the series $a_0, a_1, \ldots, a_{k-1}$.

For this, we observe that equation (\ref{eq:F}) is equivalent to
\begin{equation*}
(1-qz) \left(1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)\right)
(a_0 + a_1z +a_2z^2 + \ldots) = 1-(qz)^\gamma.
\end{equation*}
Equating the terms of same degree on both side of the equal signs leads
to the following relationship between the coefficients of the series:
\begin{equation}
\label{eq:PE0}
a_k = 
\begin{cases}
1            &\quad\text{if } k < \gamma, \\
1 -pq^\gamma &\quad\text{if } k = \gamma, \\
a_{k-1} -pq^\gamma \cdot a_{k-\gamma-1} &\quad\text{otherwise.}
\end{cases}
\end{equation}

The term of interest can be computed recursively using expression
(\ref{eq:PE0}). This approach is very efficient because every iteration
involves at most one multiplication and one subtraction. Also, the default
floating-point arithmetic on modern computers gives sufficient precision
to not worry about numeric instability for the problems considered here
(we rarely need to compute those probabilities for reads above 500
nucleotides).

This example shows how the symbolic approach gives a non-trivial and yet
very simple method to compute the probability that a read of size $k$ does
not contain an on-target exact seed.

\subsection{Example 2: on-target skip seeds}

Another example that will turn out to be essential is to compute the
probability that a read contains no on-target skip seed. We tackle the
question with the same strategy as in the previous example, so we first
recode the reads using a more useful alphabet to solve this problem.

Once again, we need to know whether a nucleotide is a sequencing error,
but this time we also need to know its phase in the repeated cycles of
skipped positions. For this, we define the skip-$n$ alphabet
$\mathcal{A}_n = \{\square, |, \Downarrow_0, \Downarrow_1, \ldots,
\Downarrow_n \}$. As before, the symbol $\square$ represents a correct
nucleotide and the symbol $|$ is a terminator appended at the end of the
read. The symbols $\Downarrow_m$ ($m \in \mathbb{N}$) represent sequencing
errors and $m$ indicates the number of nucleotides till the next
non-skipped position (\textit{i.e.} $m=0$ for nucleotides immediately
before a non-skipped position and $m=n$ for nucleotides at a non-skipped
position).

The segments in this alphabet are defined as sequences of 0 or more
symbols $\square$ followed by any of the symbols $\Downarrow_m$. The tail
is again defined as a sequence of 0 or more symbols $\square$ followed by
the symbol $|$. Given that this decomposition is unique, we can again view
a read as a sequence of segments with a tail. The example of
Fig.~\ref{fig:simple} is showed again in Fig.~\ref{fig:skip}, where the
segments in the mismatch alphabet have been replaced by the segments in
the skip-3 alphabet,

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_skip.pdf}
\caption{\textbf{The skip encoding.}
Legend.}
\label{fig:skip}
\end{figure}

The probability of occurrence of a sequencing error is $p$ so the weighted
generating function of all the symbols $\Downarrow_m$ are identical and
equal to $pz$. The weighted generating function of the symbol $\square$ is
again $qz$, and so the weighted generating function of a segment with $m$
symbols $\square$ is again $(qz)^mpz$, and the weighted generating
function of a tail segment with $m$ symbols $\square$ is again $(qz)^m$.

Reads that do not contain any on-target skip seed can still contain
segments with $\gamma$ or more symbols $\square$, so this case is more
complex than the previous one. For instance, if there is a sequencing
error in phase $r$, the status of the next $r$ nucleotides does not have
any influence on the presence of an on-target seed. Indeed, there is no
on-target seed upstream of the error by construction, and the next seed is
scheduled to start $r$ nucleotides downstream of it, so the following
segment or tail can have up to $\gamma+r-1$ symbols $\square$. So in order
to enforce the absence of on-target seed, we have to adjust the maximum
size of the segments depending on the preceding terminator.

In the skip-$n$ alphabet, reads are composed of $n+2$ different kinds of
objects: the segments terminated by the $n+1$ symbols $\Downarrow_0,
\Downarrow_1, \ldots, \Downarrow_n$, plus the tail. The dimensions of the
transfer matrix are thus $(n+2) \times (n+2)$.

As mentioned above, after the $\Downarrow_r$ terminator, we can append
segments with up to $\gamma+r-1$ symbols $\square$. The terminators of
those $\gamma+r$ possible segments are distributed according to the laws
of the modulo $n+1$ arithmetic. For instance, if the next segment has no
symbol $\square$ then it is terminated by $\Downarrow_{r+1}$, but also if
it has $n+1$ symbols $\square$ (assuming $n+1 < \gamma$). Specifying the
transfer matrix thus involves a substantial amount of bookkeeping.

\begin{equation*}
\begin{blockarray}{cccccc}
   & \Dn{0} & \Dn{1} & \ldots & \Dn{n} & \nd \\
\begin{block}{c[ccccc]}
\Dn{0} & H_{0,0}(z) & H_{0,1}(z) & \ldots & H_{0,n}(z) & J_0(z) \\
\Dn{1} & H_{1,0}(z) & H_{1,1}(z) & \ldots & H_{1,n}(z) & J_1(z) \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\Dn{n} & H_{n,0}(z) & H_{n,1}(z) & \ldots & H_{n,n}(z) & J_n(z) \\
\nd & 0 & 0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray}
\end{equation*}

\begin{eqnarray}
H_{r,s}(z) &=& pz \cdot (qz)^x \cdot \big( 1 + (qz)^{n+1} +
  \ldots + (qz)^{m(n+1)} \big), \\
\notag
  &\;& \text{where } x = r-s-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma+r-1-x}{n+1} \right\rfloor, \\
J_r(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+r-1}.
\end{eqnarray}

The transfer matrix is now too complex to compute a close expression
for $M(z)\cdot(I-M(z))^{-1}$. We return to the definition $M(z) + M(z)^2 +
\ldots$ and we observe that the terms $M(z)^{k+2}, M(z)^{k+3}, \ldots$
have no influence on the coefficients $a_0, a_1, \ldots, a_k$.

Indeed, a read of size $k$ has at most $k$ segments (plus one obligatory
tail). Since $M(z)^{s+1}$ contains the weighted generating functions of
reads with exactly $s$ segments and one tail, $a_k$ cannot depend on
$M(z)^{k+2}, M(z)^{k+3}$. More formally, we can prove by induction that
all the entries of $M(z)^k$ are divisible by $z^{k-1}$, showing that the
contribution of $M(z)^{k+2} + M(z)^{k+3} + \ldots$ to $a_0 + a_1z + \ldots
+a_kz^k$ is $0$.

So we can compute the matrix $M(z) + M(z)^2 + \ldots + M(z)^{k+1}$,
extract the entry of interest and then compute the terms of the Taylor
expansion up to order $k$. But we can do better. Since we are only
interested in the coefficients up to order $k$, we can perform all
algebraic operation on truncated polynomials of order $k$, \textit{i.e.},
we discard the coefficients of order $k+1$ or greater when multiplying two
polynomials.

But we can do even better. A read with $s$ segments contains $s$ errors.
Since read errors typically have a small probability of occurrence, all
the coefficients of $M(z)^{s+1}$ rapidly converge to $0$ as $s$ increases.
So instead of computing the matrix $M(z) + M(z)^2 + \ldots + M(z)^{k+1}$,
we can interrupt the summation after a certain power of $M(z)$ because the
terms become negligible.

\begin{equation}
\label{eq:bound}
Pr(X \geq s) \leq \exp \left( (s-k)\log \frac{k-s}{k(1-p)} -s\log
\frac{s}{kp} \right).
\end{equation}

Using the formula above, we can thus bound the probability that a read has
$s$ segments or more. We compute the matrices $M(z) + M(z)^2 + \ldots
+M(z)^{s+1}$ where the weighted generating functions have been replaced by
truncated polynomials and we extract the entry at the first column and the
last row. When the bound is lower than a set fraction $\varepsilon$ of the
smallest coefficient, we stop the computations. Typically $\varepsilon =
0.01$ so this method ensures that the probabilities that a read of size
$k$ has no good seeds are accurate to within 1\%.

\section{Exact seeds}

\subsection{Corollary: this title sucks}

The results of this section allow us to derive the probabilities of false
positives when using seeds of fixed size $\gamma$.

The complementary of the event $E_0$ is that the read contains at least
one match of size $\geq \gamma$ for the target, implying that there is a
seed for the target\footnote{This is the major difference with MEM seeds,
where the logical implication does not hold.}. With our assumptions this
means that the read is mapped to the correct location. The probability of
outcome $i.$, a true positive, is thus $1-P(E_0)$.

The event $E_0 \cap \ldots \cap E_N$ is that the read contains no match of
size $\geq \gamma$ for either the targer or any of the off-target
sequences. In other words, the probability of outome $iii.$, a false
negative, is $P(E_0 \cap \ldots \cap E_N)$.

The event $E_0 \cap (E_1 \cap \ldots \cap E_N)^c$ is that the read
contains no match of size $\geq \gamma$ for the target, but contains a
match of size $\geq \gamma$ for at least one of the off-target sequences.
In other words, the probability of outcome $ii.$, a false positive, is
$P(E_0) - P(E_0 \cap \ldots \cap E_N)$.

It is easy to see that the three probabilities sum to 1, as expected.
Below is a summary of the probabilities of interest.

\begin{eqnarray}
P(i.) &=& 1-P(E_0) \\
P(ii.) &=& P(E_0) \cdot \left( 1- \left( \frac{P(E_0
  \cap E_1)}{P(E_0)} \right)^N\right) \\
P(iii.) &=& P(E_0) \cdot \left( \frac{P(E_0 \cap E_1)}{P(E_0)} \right)^N
\end{eqnarray}

In the expressions above, we have used equation (\ref{eq:PE0E1}) to
highlight that these probabilities only depend on $P(E_0)$ and $P(E_0 \cap
E_1)$, which can be computed precisely as shown in sections~\ref{sec:cPE0}
and \ref{sec:cPE0E1}.

Note that the formulas also hold when the target has no duplicate
sequence. In this case, only outcome $i.$ and $iii.$ are possible, but
since $N=0$ we see that the second probability is 0, as expected.

\section{Skip seeds}
